---
title: "Spatial uncertainty propagation analysis"
author: "Kasia Sawicka and Gerard Heuvelink"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
# output: rmarkdown::word_document
subtitle: Case study with calling external model.
vignette: >
  %\VignetteIndexEntry{Spatial uncertainty propagation analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE}
knitr::opts_chunk$set(
    comment = NA,
    quiet = TRUE,
    progress = FALSE,
    tidy = FALSE,
    cache = FALSE,
    message = FALSE,
    error = FALSE, # FALSE: always stop execution.
    warning = TRUE,
    dpi = 100
)
```

```{r, echo = FALSE}
knitr::opts_knit$set(global.par = TRUE)
```

```{r, echo = FALSE}
par(mar = c(3, 3, 2, 2), mgp = c(1.7, 0.5, 0), las = 1, cex.main = 1, tcl = -0.2, cex.axis = 0.8,
    cex.lab = 0.8)
```


### Introduction

Ofter, environmental models are developed in other languages than R, for example C or FROTRAN. It significanly speeds up processing. In this simple example, we show how to perform uncertainty analysis with a basic model written in C.

<br>

### Adapted Monte Carlo methodology for uncertainty analysis

The adapted uncertainty propagation analysis approach is based on the Monte Carlo method that computes the output of the model repeatedly, with input values that are randomly sampled from their pdfs. The set of model outputs forms a random sample from the output pdf. The method thus consists of the following steps:

1.	Characterise uncertain model inputs/parameters with pdfs.
1.	Repeatedly sample from the pdfs of uncertain inputs/parameters.
1.	Run model with sampled inputs and store model outputs.
1. 	Compute summary statistics of model outputs.

<br>

### Uncertainty propagation analysis with 'spup'

```{r}
library(magrittr) # piping
library(whisker)  # templating
library(readr)    # fast I/O (huge files)
library(purrr)    # functional programming tools
library(dplyr)    # a grammar of data manipulation

setwd("D:/GD/Wageningen/QUICS/RPackage/spup/vignettes")
```


When using external models we need additional code to

1. modify ASCII input files
1. run the external model


## Modify ASCII input files (rendering)

For rendering ASCII input files, we use the mustache templating framework (https://mustache.github.io). In R this is implemented in the package `whisker`.

First we define a 'container' class to store all templates with model inputs. The aim of this class is to

1. organise model input files;
1. perform some checks


The constructor of this class is given below:

```{r}
template <- function(filenames) {
    filenames %>% 
        walk(
            function(x) {
                if (!grepl(pattern = "\\.template$", x = x)) {
                    stop(
                        "File %s does not have extension 'template'" %>% 
                            sprintf(x %>% sQuote), 
                        call. = FALSE
                    )
                }
            }
        )
    filenames %>% 
        walk(
            function(x) {
                if (!file.exists(x)) {
                    stop(
                        "File %s not found" %>% 
                            sprintf(x %>% sQuote), 
                        call. = FALSE
                    )
                }
            }
        )
    class(filenames) <- "template"
    filenames
}

print.template <- function(x, ...) {
    cat("container containing the following template(s):\n")
    print(as.character(x))
}
```



A template is simply a model input file with:

1. the additional extension `.template`. 
1. input that needs to be modified is replaced by mustache-style tags. 


For example, suppose a model needs the input file: `input.txt`. Suppose further that this input file contains two parameters b0 and b1. The contents of the original file may look like:

```{r}
read_lines("examples/input.txt")
```

where the first number is `b0` and the second number is `b1`.

Then, the corresponding template file should have filename `input.txt.template`. It contains:

```{r}
read_lines("examples/input.txt.template")
```

So, the original numbers are replaced by symbols b0 and b1 placed in moustaches `{{` and `}}`.




Rendering is the process of replacing the tags in moustaches by text. For this, we provide a set of render-methods.

```{r}

render <- function(x, ...) {
    UseMethod("render")
}

render.character <- function(x, ...) {
    dots <- list(...)
    if (length(dots) == 0L) {
        return(x)
    }
    x %>% 
        whisker.render(dots)
}

render.template <- function(x, ...) {
    sub(pattern = "\\.template$", replacement = "", x = x) %T>% 
        walk(function(x) {
            readLines(paste(x, "template", sep = "."), warn = FALSE) %>% 
            render(...) %>%
            writeLines(x)
        }) %>%
        as.character
}

```



For example:
```{r}
my_template <- "Hello {{name}}. How are you doing?"

my_template %>% 
    render(name = "Kasia")

```

This example calls `render.character`. It is also possible to fill out an entire table:

```{r}
my_template <- c(
    "| x | y |",
    "|---|---|",
    "{{#MY_TABLE}}",
    "| {{X}} | {{Y}} |",
    "{{/MY_TABLE}}"
)

my_table <- data.frame(X = 1:5, Y = letters[1:5])
my_table

my_template %>% 
    render(MY_TABLE = unname(rowSplit(my_table))) %>%
    cat
```

See the `whisker`-package (or https://mustache.github.io) for more information on the syntax. 


A template stored as a file will always be rendered on disk. Let's return to our template:

```{r}
my_template <- template("examples/input.txt.template")
```

with contents

```{r}
my_template %>% 
    read_lines
```


Rendering will create a new file, called `input.txt`.

```{r}
my_template %>% 
    render(b0 = 3, b1 = 4)
```

As can be seen above, the path of this file is also the return value of the render method. This facilitates further processing by means of the pipe-operator:

```{r}
my_template %>% 
    render(b0 = 3, b1 = 4) %>%
    read_lines
```



## Running external models

An external model can always be called from R by means of the `system` or `system2` function. To facilitate this, the following wrapper function comes in handy:

```{r}
executable <-
function(filename) {
    if (!file.exists(filename)) {
        stop(
            "%s not found" %>% sprintf(sQuote(filename)), 
            call. = FALSE
        )
    }
    function() {
        root <- dirname(filename)
        owd <- setwd(root)
        on.exit(setwd(owd))
        system2(basename(filename), wait = TRUE)
    }
}
```


Suppose we got an external model written in the C language: `dummy_model.C`. The executable on MS-Windows is called `dummy_model.exe`. We can now wrap this model as follows:

```{r}
dummy_model <- executable("examples/dummy_model.exe")
```


Running this model gives
```{r}
# create template
my_template <- template("examples/input.txt.template")

# render the template
render(my_template, b0 = 3.1, b1 = 4.2)

# run external model
dummy_model()

# read output (output file of kasia-model is "output.txt")
scan(file = "examples/output.txt", quiet = TRUE)
```


An example of multiple realizations:
 
 1. render template
 1. run model
 1. read results
 1. process results

```{r}

n_realizations <- 100

n_realizations %>%
    rerun({
        # render template
        render(my_template, b0 = rnorm(n = 1), b1 = runif(n = 1))
        
        # run model
        dummy_model()

        # read output
        scan("examples/output.txt", quiet = TRUE)
    }) %>%
    set_names(paste0("r", 1:n_realizations)) %>% 
    as_data_frame %>%
    apply(MARGIN = 1, FUN = quantile)    
```




### Acknowledgements

We thank Dennis Walvoort for his valuable contribution to the development of this package.

This project has received funding from the European Union’s Seventh Framework Programme for research, technological development and demonstration under grant agreement no 607000.

